---
title: "Danish Kings"
author: "Marie W. Schmidt"
date: "2026-03-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


The task here is to load your Danish Monarchs csv into R using the `tidyverse` toolkit, calculate and explore the kings' duration of reign with pipes `%>%` in `dplyr` and plot it over time.

## Load the kings

Make sure to first create an  `.Rproj` workspace with a `data/` folder where you place either your own dataset or the provided `kings.csv` dataset. 

1. Look at the dataset that are you loading and check what its columns are separated by? (hint: open it in plain text editor to see)

The columns in the dataset is seperated by ´;´  

2. Create a `kings` object in R with the different functions below and inspect the different outputs.
-  `read.csv()`
-  `read_csv()`
-  `read.csv2()` 
-  `read_csv2()`
```{r eval=TRUE}
library(tidyverse)
# FILL IN THE CODE BELOW and review the outputs

kings1 <- read.csv("data/danish_kings.csv")

kings2 <- read_csv("data/danish_kings.csv")

kings3 <- read.csv2("data/danish_kings.csv")

kings4 <- read_csv2("data/danish_kings.csv")
```

1. Which of these functions is a `tidyverse` function? Read data with it below into a `kings` object

The read_csv and read_csv2 is a tidyverse function. They read data as a tibble, whereas read.csv reads the data as   a date.frame. This is also the reason for why kings1 and kings3 above wont work with our dataset. Tidyverse overrules the default setting. Because my data-columns are seperated by comma, I need to use the the read_csv2 function, and kings4 above is therefore the correct solution for my dataset.
  
2. What is the result of running `class()` on the `kings` object created with a tidyverse function.

The answer: [1] "spec_tbl_df" "tbl_df"      "tbl"         "data.frame" tells us, that the object is a tibble.
  
3. How many columns does the object have when created with these different functions?

The object has 7 columns
  
4. Show the dataset so that we can see how R interprets each column

```{r eval=TRUE}
#1 
kings<-read_csv2("data/danish_kings.csv")

#2
class(kings)

#3
ncol(kings)

#4
glimpse(kings)

```


## Calculate the duration of reign for all the kings in your table
You can calculate the duration of reign in years with `mutate` function by subtracting the equivalents of your `startReign` from `endReign` columns and writing the result to a new column called `duration`. But first you need to check a few things:

* Is your data messy? Fix it before re-importing to R
* Do your start and end of reign columns contain NAs? Choose the right strategy to deal with them: `na.omit()`, `na.rm=TRUE`, `!is.na()`


Create a new column called  `duration` in the kings dataset, utilizing the `mutate()` function from tidyverse. Check with your group to brainstorm the options.

```{r eval=TRUE}
# My dataset already included a column for duration, but for the purpose of this assignment I created a new one
kings_duration <- kings %>% 
  filter(!is.na(start_of_reign&end_of_reign)) %>% 
  mutate(duration=end_of_reign-start_of_reign)

```

## Calculate the average duration of reign for all rulers
Do you remember how to calculate an average on a vector object? If not, review the last two lessons and remember that a column is basically a vector. So you need to subset your `kings` dataset to the `duration` column. If you subset it as a vector you can calculate average on it with `mean()` base-R function. If you subset it as a tibble, you can calculate average on it with `summarize()` tidyverse function. Try both ways!

* You first need to know how to select the relevant `duration` column. What are your options?
* Is your selected `duration` column a tibble or a vector? The `mean()` function can only be run on a vector. The `summarize()` function works on a tibble.
* Are you getting an error that there are characters in your column? Coerce your data to numbers with `as.numeric()`.
* Remember to handle NAs: `mean(X, na.rm=TRUE)`


```{r eval=TRUE}
kings_duration%>% 
  summarise(mean_duration=mean(duration,na.rm = TRUE))

```
Answer: the average duration of reign is 20.10204 years.

## How many and which kings enjoyed a longer-than-average duration of reign?

You have calculated the average duration above. Use it now to `filter()` the `duration` column in  `kings` dataset. Display the result and also  count the resulting rows with `count()`

```{r eval=TRUE}
kings_duration %>% 
  filter(duration>20.10204) 

kings_duration %>% 
  filter(duration>20.10204) %>%
  count()

```
Answer: 22 kings reign was above the average duration.

## How many days did the three longest-ruling monarchs rule? 

* Sort kings by reign `duration` in the descending order. Select the three longest-ruling monarchs with the `slice()` function
* Use `mutate()` to create `Days` column where you calculate the total number of days they ruled
* BONUS: consider the transition year (with 366 days) in your calculation!

```{r}
#1
longest_reign <- kings_duration %>% 
  arrange(desc(duration)) %>% 
  slice(1:3)

#2
longest_reign %>% 
  mutate(duration_days=duration*365)

```


## Challenge: Plot the kings' duration of reign through time

What is the long-term trend in the duration of reign among Danish monarchs? How does it relate to the historical violence trends ? 

* Try to plot the duration of reign column in `ggplot` with `geom_point()` and  `geom_smooth()`
* In order to peg the duration (which is between 1-99) somewhere to the x axis with individual centuries, I recommend creating a new column `midyear` by adding to `startYear` the product of `endYear` minus the `startYear` divided by two (`startYear + (endYear-startYear)/2`).
* Now you can plot the kings dataset, plotting `midyear` along the x axis and `duration` along y axis
* BONUS: add a title, nice axis labels to the plot and make the theme B&W and font bigger to make it nice and legible!

```{r}
kings_trend <- kings_duration %>% 
  mutate(midyear=start_of_reign+(end_of_reign-start_of_reign)/2)

kings_trend %>% 
  ggplot(aes(x=midyear,y=duration))+
  geom_point()+
  geom_smooth()+
  labs(title = "Danish kings: Duration of reign",
       x="Midyear",
       y="Duration")+
  theme_bw()+
  theme(text = element_text(size = 16))

```

And to submit this rmarkdown, knit it into html. But first, clean up the code chunks, adjust the date, rename the author and change the `eval=FALSE` flag to `eval=TRUE` so your script actually generates an output.
Well done!


